SMT-Based Generation of Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-03-16 2019-ActaInformatica/ActaInformatica.tex
changes in writing

\input{XFIG/macros.tex}         % Definitions of macros used in XFig figures
\usepackage{macrospNets}

@@

\title{Using an SMT engine to generate Symbolic Automata%
%
  \thanks{This work was partially funded by the Associated Team FM4CPS
    between INRIA and ECNU, Shanghai}%
}

@@

\abstract{Open pNets are used to model the behaviour of open systems, both
synchronous or asynchronous, expressed in various calculi or
languages. They are endowed with a symbolic operational semantics in
terms of so-called ``Open Automata''. This allows us to check properties of
such systems in a compositional manner. 
We implement an algorithm computing these semantics,
building predicates expressing the synchronization conditions
between the events of the pNet sub-systems. Checking such
predicates requires symbolic reasoning over first order logics, but
also over application-specific data. We use the Z3 SMT engine to check
satisfiability of the predicates, and prune the open automaton of its
unsatisfiable transitions.
As an industrial oriented use-case, we use so-called ``architectures''
for BIP systems, that have been used in the framework of an ESA project and to specify the control software of
a nanosatellite at the EPFL Space Engineering Center. We use pNets to encode a BIP architecture
extended with explicit data, and compute its
open automaton semantics. This automaton may be used to prove
behavioural properties; we give 2 examples, a safety and a liveness
property.
\TODO{Add new contrib}
\todoMargin{Should we remove Min ?}
\keywords{Networks of Synchronised Automata, Symbolic Behavioral Semantics, Compositional Analysis of Software Components, SMT}
}

@@

In the nineties, several 
works extended the basic behavioural models based on labelled
transition systems to address value-passing or parameterized systems, using
various symbolic encodings of the
transitions~\cite{deSimone85,Larsen87,HennessyLin:TCS95,Linconcur96}. 
In \cite{Linconcur96}, H.M. Lin addressed value-passing calculi, for which he
developed a symbolic behavioural semantics, and proved algebraic properties.
Separately J. Rathke~\cite{HennessyRathke:TCS98} defined another
symbolic semantics for 
a parameterized broadcast calculus, together with strong and weak bisimulation
equivalences, and developed a symbolic model-checker based on a tableau
method for these processes. Thirty years later, no
practical verification approach and no verification platform are
using this kind of approaches to provide proof methods for
value-passing processes or open process expressions.

Parameterized Networks of Synchronized Automata (pNets) were proposed
to give a behavioural specification formalism for distributed
systems, synchronous, asynchronous, or heterogeneous. They are used in
VerCors~\cite{HKLM:Foclasa14}, a platform for designing and 
verifying distributed systems, as the intermediate language for various
high-level languages. The high-level languages in VerCors formalize
each component of the distributed system and their composition.
pNets provides the core low-level semantic formalism for VerCors, and
is made of a hierarchical composition of (value-passing) automata,
called parameterized labelled transition systems (pLTS), where each
hierarchical level defines the possible synchronization of the lower levels.
Traditionally, pNets have been used to formalize fully
defined systems or softwares. But we want also to define and reason
about incompletely defined systems, like program skeletons, operators,
or open expressions of process calculi.
The open pNet model addresses this problem, using
``holes'' as process parameters, representing \emph{unspecified} subsystems.
The pNet model was developed in a series of
papers~\cite{HMZ:PDP15,henrio:Forte2016} in which many examples have been
introduced showing its ability to encode the operators from some
other algebras or  program skeletons.
The operational semantics of an (open) pNet is defined as an
Open Automaton in which Open Transitions contain logical predicates
expressing the relations between the behaviour of the holes, and the
global behaviour of the system. In the previous publication,
only a sketch of a procedure allowing to compute these semantics was
presented, together with a proof of finiteness of the open automaton, under
reasonable hypotheses on the pNet structure.

Implementing these semantics raised several challenges, in order:
\begin{itemize}
  \item to get a tool that could be applied to pNets representing
    various languages, in particular various actions algebras,
    with their specific decision theories,
  \item to separate clearly the 
      algorithm generating the transitions of the open automaton from
      combination of all possible (symbolic) behaviours, from
      the symbolic reasoning part, specifically here using an SMT
      engine to check the
      satisfiability of the predicates generated by our algorithm,
  \item to build a prototype and validate the approach on our basic
    case-studies, and understand the efficiency of the interaction
    with the SMT solver.
\end{itemize}

In the long term, we want to be able to check the equivalence between
open systems encoded as pNets. The equivalence between pNets is
``FH-bisimulation'' \cite{henrio:Forte2016}, a dedicated version of
symbolic bisimulation taking the  
predicate of the open transitions into account when matching such
open transitions. We foresee that the interplay with the SMT solver
that we use here for satisfiability of open transitions will be
similar with what we need when proving (symbolic) equivalence between open
transitions.

\paragraph{Contribution}
In the article we show how:
\begin{itemize}
\item We define the open automaton generation algorithm. We
  implemented a full working prototype, within the 
  VerCors platform. In the process, we improved the
  semantics rules from~\cite{henrio:Forte2016}, and add features in
  the algorithm to deal with the full 
  model, including management of variables and assignments.
\item We implement the interaction between our algorithm and the Z3
    SMT solver, for checking satisfiability of the transitions
    generated by the algorithm.
    \item \ERIC{we define a strategy checking satisfiability of open transitions in the course the residual algorithm, and show that this ensures that we still generate the reachable part of the open automaton, and that what is generated is minimal with respect to the decidability of predicate satisfiability} 
\item We show the interest of this approach \ERIC{on various use-cases from different settings}, including an
    industry-inspired case-study, namely one architectural pattern
    extracted (and extended) from the BIP specification of a nanosatellite on-board
    software.
\end{itemize}

@@

%% There is not much research work towards trying 
Very few attempts were made to develop symbolic bisimulation approaches
for the value-passing process algebras and languages\mdash our long-term goals\mdash
especially, there is no algorithmic treatment of the symbolic systems developed by
interacting with automatic theorem provers. The closest work is the
one already mentioned from J. Rathke~\cite{HennessyRathke:TCS98},
who developed the symbolic bisimulation for a
calculus of broadcasting system (CBS). CBS is similar with classic
process calculi such as CCS and CSP, but communicating by broadcasting
values, %one-to-many communication instead of one-to-one communication and
transmitting values without blocking. That makes the definition of
the symbolic semantic and bisimulation equivalence different from the
classic works.

For other applications, such as the analysis of 
programming languages, there exist dedicated platforms using
external automatic theorem 
provers (ATP) or automatic tactics from interactive theorem
provers (ITP), to perform symbolic reasoning, and for example to
discharge some subgoals in the proofs.
Tools like Rodin~\cite{deharbe2013,deharbe2014} have
already integrated several provers, like Z3, as modules for proving
the proof obligations generated from a user model. 
The prover we use, which also happens to be Z3, is developed by Microsoft Research
based on the satisfiability modulo 
theories framework (SMT), is mainly applied in extended static checking, test case
generation, and predicate abstraction.
In a similar way, there are several ATPs/ITPs we could consider to use for
result pruning and bisimulation checking in our algorithm, as an
alternative to Z3, such as CVC4~\cite{barrett:CAV2011},
Coq~\cite{armand:CPP2011} or others.

@@

\paragraph{Structure.}
In section
\ref{section:pnets} we give a description and a formal definition of
the pNet model, as found in previous publications. Then in section
\ref{section:examples} we present our use-case, based on a BIP
architecture from the nano-satellite case-study. 
Section \ref{section:op-semantics} recalls briefly the operational semantics
of pNet.
Section \ref{section:implementation} explains in details the algorithm
used to compute this semantics, including the interaction with Z3, and
shows the full result of the semantic computation on the running example.
Finally we conclude and discuss perspectives in Section
\ref{section:conclusion}.

@@

This section introduces pNets and the notations we will use in
this paper. Then it gives the formal definition of pNet structures,
together with an operational semantics for open pNets.

pNets are tree-like structures, where the leaves are either
\emph{parameterized labelled transition systems (pLTSs)}, expressing the
behaviour of basic processes, or \emph{holes}, used as placeholders
for unknown processes, of which we only specify their set of possible
actions, named \emph{sort}.
Nodes of the tree (pNet nodes) are synchronizing artifacts, using a
set of \emph{synchronization vectors} that express the possible
synchronization between the parameterized actions of a subset of the
sub-trees.

%\smallskip\noindent
\paragraph*{Notations.}
We extensively use indexed structures
over some countable indexed sets, which are equivalent to mappings over
the countable set. % . The indexes will usually be
% integers, bounded or not. Such an indexed family is
%denoted
%follows:
$a_i^{i\in I}$
denotes a family of elements $a_i$ indexed over the
set $I$.
When this is not
ambiguous, we shall use notations for sets, and typically write
``indexed set over I'' when formally we should speak of multisets, and
write $x\in a_i^{i\in I}$ to mean $\exists i\in I.\, x=a_i$.  An empty
family is denoted $\emptyset$. We
denote classically $\overline{a}$ a family when the indexing set is
irrelevant.  $\uplus$ is the disjoint union on indexed sets.

\paragraph*{Term algebra.}
Our models rely on a notion of parameterized actions that are
symbolic expressions using data types and variables. As we want to encode
the low-level behaviour of possibly very different
programming languages, we do not want to impose one specific algebra
for denoting actions, nor any specific communication mechanism. So we
leave unspecified the constructors of the algebra that will allow building
expressions and actions. Moreover, we use a generic {\em action interaction}
mechanism, based on unification between two or more action
expressions. This will be used in the semantics of synchronization
vectors to express various kinds of communication or synchronization mechanisms.

@@

A pLTS is a labelled transition system with variables, which can be
manipulated, defined, or accessed inside states, actions, guards, and
assignments. 
%
%Without loss of generality and to simplify the formalization, we suppose 
%here that variables are local to 
%state: each state has its set of variables disjoint from the others. Transmitting 
%variable values from one state to the other can be done by explicit assignment. 
%
Each state has its set of variables called \emph{state variables}, 
which can only be modified by the assignment in transitions targeting this state. 
A global state variable of a pLTS is a state variable defined in all states.
% $x$ satisfying $\forall s\!\in\! S .\, x\!\in\! \vars(s)$. 
%
%Similarly, to simplify the management of variables and without loss of expressivity, we 
%suppose that transitions looping to the same state does not do assignments.
Note that we make no assumptions on finiteness of the set of states, nor
on finite branching of the transition relation.

We first define the set of actions a pLTS can use.  Let $a$
range over action labels, $\symb{op}$ are operators, and $x_i$ range over
variable names. Action terms are:
\[
\begin{array}[l]{rcl@{\quad}p{5.5cm}}
  \alpha\in\AlgA&::=&a(p_1,\ldots,p_n)&\text{action terms}\\
  p_i&::=& ~\symb{Expr}&\text{parameters}\\
  \symb{Expr}&::=& \symb{Value}~|~x~|~\symb{op}(\symb{Expr}_1,\dots,\symb{Expr}_n)&\text{expressions}
\end{array}
\]
%\[
%\begin{array}[l]{rcl@{\quad}p{5.5cm}}
%  \alpha\in\AlgA&::=&a(p_1,\ldots,p_n)&\text{action terms}\\
%  p_i&::=& ?x~|~\symb{Expr}&\text{parameters (input variable or expression)}\\
%  \symb{Expr}&::=& \symb{Value}~|~x~|~\symb{op}(\symb{Expr}_1,..,\symb{Expr}_n)&\text{Expressions}
%\end{array}
%\]

\begin{definition}[pLTS]
\label{pLTS}
%\noteSB{It's a bit strange that algebra presentations are not referred to here}
Given a term algebra $\Talg$, a pLTS is a tuple
$pLTS\triangleq\mylangle S,s_0, \to\myrangle$ where:
\begin{itemize}
\item[$\bullet$]
$S$ is a set of \emph{states}, with $s_0 \in S$ the \emph{initial state}.
\item[$\bullet$] $\to \subseteq S \times L \times S$ is the \emph{transition relation}, with
$L$ the set of labels of the form
$\langle \alpha,~e_b,~(x_j\!:= {e}_j)^{j\in J}\rangle$,
  where $\alpha \in\AlgA_{\P}$ is a parameterized action,
    $e_b \in \AlgB_{\P}$ is a guard, and
    expressions  $\AlgE_{\P}\cup\AlgA_{P}$ are assigned to $x_j$.
If 
$s \xrightarrow{\langle \alpha,~e_b,~(x_j\!:= {e}_j)^{j\in
		J}\rangle} s'$ then 
                $\vars(e_b)\!\subseteq\! \vars(s)\cup\vars(\alpha)$, and
		$\forall j\!\in\! J .\,\vars(e_j)\!\subseteq\! \vars(s)\land 
x_j\!\in\!\vars(s')$.
%\item[$\bullet$]
%The $x_j$ are \emph{state variables} of %% the state $s'$. 
\end{itemize}
\end{definition}

Now, we define
pNet nodes as constructors for hierarchical structures.
A pNet node has a set of sub-pNets that can be either pNets or pLTSs, and a
set of \emph{holes}, playing the role of process parameters
(i.e. unknown in the environment).

@@

%% \begin{definition}[pNets]\label{def-pnets}
%% A pNet is a hierarchical structure, whereof the leaves are pLTSs and holes:
%% $\pNet::= pLTS~|~\mylangle \pNet_i^{i\in I}, S_j^{j\in J}, \symb{SV}_k^{k\in K}\myrangle$,
%% where
%% \begin{itemize}
%% \item[$\bullet$] %\noteEM{Pire que ca, $\I$ n'est pas
%%                  %defini...}\noteSB{Why is one $\I$ and the other
%%   %$\I_\P$?}
%%   $I$ and $J$ are the sets over which are indexed sub-pNets and holes, respectively, such that: $I\cap J=\emptyset$ and  $I\cup J\neq\emptyset$.

@@

%% \item[$\bullet$] $S_j \subseteq \AlgA$ is a set of action terms,
%%   denoting the $\Sort$\footnote{The definition of $Sorts$ (the
%%     set of actions of a Hole  
%%   or pNet), and also of $Leaves$ and $Holes$ (respectively all pLTSs
%%   and all holes in a pNet), can be found in \cite{henrio:Forte2016}.}
%%   of hole $j$.

@@

  \begin{definition}[pNets]\label{def-pnets}
    A pNet is a hierarchical structure where leaves are pLTSs and holes:
   %%  \footnote{Typing can 
%% be added to pNets by defining and checking their interfaces, called 
%% sort~\cite{HMZ-FORTE2016}}
$\pNet\triangleq \pLTS~|~\mylangle \pNet_i^{i\in I}, J, \symb{SV}_k^{k\in 
K}\myrangle$, with $I, J, K$ potentialy infinite,
where
\begin{itemize}
\item[$\bullet$] $\pNet_i^{i\in I}$ is the family of sub-pNets;
%  $\pNet_i^{i\in I}$ is a family of sub-pNets where $I\in\I_\P$ is the set over which 
%sub-pNets are indexed.

@@

In Fig.~\ref{schema:EnableEncodings}, we show examples of these
constructs, with two pLTSs, one hole and one pNet node encoding our running example.

@@

The semantics of open pNets will be defined as an open automaton, that
is an automaton where each transition composes transitions 
of several
% \noteSB{These should all be pLTSs, right? Or did I miss
%  something?}\noteEM{Hum, nor really, something intermediate; the actions there can be
%  parameterised actions, but the label has no guard and no
%  assignments}
LTSs with the actions of some holes; the transition occurs if some predicates hold,
and can involve a set of state modifications. Each state of an open
automaton has a set of \emph{state variables} that can be assigned by
incoming transitions. Strictly speaking, the LTSs at the leaves of the
open automaton are a restricted form of pLTSs, where labels are
parametrised actions, but include no guard nor assignments.

@@

%
%Then the semantics of a pNet is characterized by a set of {\em open
%transitions}, where the hypotheses on process parameters are
%replaced by 1) transitions of the pLTSs at the leaves, and 2) formal
%hypotheses on the transitions of the holes. A {\em predicate} is used
%to relate the parameters and names appearing in the actions of the
%leaves and the holes involved in the rules, but also appearing in  the resulting action.

@@

\paragraph{States of open pNets.}\label{def-states}
  A state of an open pNet is a tuple of the
  states of its leaves (in which we denote tuples
  in structured states as $\triangleleft\ldots\triangleright$).
  For any pNet p, let $\overline{\mathit{Leaves}} = \mylangle S_i,{s_i}_0, \to_i\myrangle^{i \in L}$ be the set of pLTS at its leaves,
  then $\mathit{States}(p) = \{\ostate{s_i^{i\in L}}
  |\, \forall i\in L. s_i \in S_i\}$.
A pLTS being its own single leave:
  $\mathit{States}(\mylangle S,s_0, \to\myrangle) = \{\ostate{s}|\, s \in S\}$.
The initial state is defined as:
$\mathit{InitState}(p) = \ostate{{{s_i}_0}^{i\in L}}$.

@@

\paragraph{Structural Semantic Rules:}
The semantics of pNet in term of open automata has been defined in
\cite{henrio:Forte2016}, in the form of 2 structural rules, one for
pLTSs, one for pNet nodes. These rules are slightly improved, adding guards in the synchronisation vectors and syntax for universal quantifier in the guards (see~\cite{Avocs-RR}).

In \cite{henrio:Forte2016} we also proved:
\begin{theorem}[Finiteness]
  \label{Th:finiteness}
  Given an open pNet $pnet=\mylangle \overline{\pNet},\overline{S}, \symb{SV}_k^{k\in K}\myrangle$ with leaves $l_i^{i\in I}$ and holes $h_j^{j\in
  J}$, if the sets $I$ and $J$ are finite, if the synchronisation vectors of all pNets 
  included in  $pnet$   are finite, and if
$\forall i \in L.\, finite{(states(l_i))} \text{ and } l_i$
has a finite number of state variables, then its semantics is an open automaton 
$\mathcal{T}$ with finitely many states and transitions.
\end{theorem}
  Remark that all the elements of such pNets and open automata being
symbolic, they can represent many classes of unbounded systems.

>>>>

\input{figures/macros.tex}         % Definitions of macros used in XFig figures
\usepackage{macrospNets}

@@

\title{SMT-Based Generation of Symbolic Automata%
%
  \thanks{This work was partially funded by the Associated Team FM4CPS
    between INRIA and ECNU, Shanghai}%
}

@@

\abstract{Open pNets are used to model the behaviour of open systems, both
synchronous and asynchronous, expressed in various process calculi or
languages. They are endowed with a symbolic operational semantics in
terms of open automata, which allows us to check properties of
such systems in a compositional manner. 
We implement an algorithm computing these semantics,
building predicates expressing the synchronization conditions
between the events of the pNet sub-systems. Checking such
predicates requires symbolic reasoning about first order logics and application-specific data. We use the Z3 SMT engine to check
satisfiability of the predicates.
%As an industrial oriented use-case, %we use so-called ``architectures''
%for BIP systems, that have been used in the framework of an ESA project and to specify the control software of a nanosatellite at the EPFL Space Engineering Center. 
As a use-case, we use pNets to encode a BIP architecture
extended with explicit data, and compute its
open automaton semantics. This automaton may be used to prove
behavioural properties such as safety and liveness.
\TODO{Add new contrib}
\todoMargin{Should we remove Min ?}
\keywords{pNets, software components, symbolic semantics, SMT}
}

@@

In the nineties, several 
works extended the basic behavioural models based on labelled
transition systems to formalize value-passing or parameterized systems, using
various symbolic encodings of the
transitions~\cite{deSimone85,Larsen87,HennessyLin:TCS95,Linconcur96}. 
In \cite{Linconcur96}, Lin considered value-passing calculi, for which he
developed a symbolic behavioural semantics, and proved algebraic properties.
Separately, Rathke~\cite{HennessyRathke:TCS98} defined another
symbolic semantics for 
a parameterized broadcast calculus, together with strong and weak bisimulation
equivalences, and developed a symbolic model-checker based on a tableau
method for these processes. Thirty years later, no
practical verification approach and no verification platform use this kind of approaches to provide proof methods for
value-passing processes or open process expressions.

Parameterized Networks of Synchronized Automata (pNets) were proposed
to give a behavioural specification formalism for distributed
systems, synchronous, asynchronous, or heterogeneous. They are used in
VerCors~\cite{HKLM:Foclasa14}, a platform for designing and 
verifying distributed systems, as the intermediate language for various
high-level languages. The high-level languages in VerCors formalize
each component of the distributed system and their composition.
pNets provides the core low-level semantic formalism for VerCors, and
is made of a hierarchical composition of (value-passing) automata,
called parameterized labelled transition systems (PLTS), where each
hierarchical level defines the possible synchronization of the lower levels.
Traditionally, pNets have been used to formalize fully
defined systems or softwares. But we want also to define and reason
about incompletely defined systems, like program skeletons, operators,
or open expressions of process calculi.
The open pNet model addresses this problem, using
``holes'' as process parameters, representing \emph{unspecified} subsystems.
The pNet model was developed in a series of
papers~\cite{HMZ:PDP15,henrio:Forte2016} in which many examples have been
introduced showing its ability to encode the operators from some
other algebras or  program skeletons.
The operational semantics of an open pNet is defined as an
open automaton in which open transitions contain logical predicates
expressing the relations between the behaviour of the holes, and the
global behaviour of the system. In the previous publication,
only a sketch of a procedure allowing to compute these semantics was
presented, together with a proof of finiteness of the open automaton, under
reasonable hypotheses on the pNet structure.

Implementing these semantics raised several challenges, in order:
\begin{itemize}
  \item to get a tool that could be applied to pNets representing
    various languages, in particular various action algebras,
    with their specific decision theories,
  \item to separate clearly the 
      algorithm generating the transitions of the open automaton from
      combination of all possible (symbolic) behaviours, from
      the symbolic reasoning part, specifically here using an SMT
      engine to check the
      satisfiability of the predicates generated by our algorithm,
  \item to build a prototype and validate the approach on our basic
    case-studies, and understand the efficiency of the interaction
    with the SMT solver.
\end{itemize}

In the long term, we would like to check the equivalence between
open systems encoded as pNets. The equivalence between pNets is
``FH-bisimulation'' \cite{henrio:Forte2016}, a dedicated version of
symbolic bisimulation taking the  
predicate of the open transitions into account when matching such
open transitions. We foresee that the interplay with the SMT solver
that we use here for satisfiability of open transitions will be
similar with what we need when proving (symbolic) equivalence between open
transitions.

\paragraph{Contribution}
In the current work we contribute in the following aspects.
\begin{itemize}
\item We present an open automaton generation algorithm. We
  implement a full working prototype, within the 
  VerCors platform. In the process, we improve the
  semantics rules from~\cite{henrio:Forte2016}, and add features in
  the algorithm to deal with the full 
  model, including management of variables and assignments.
\item We implement the interaction between our algorithm and the Z3
    SMT solver, for checking satisfiability of the transitions
    generated by the algorithm.
    \item We define a strategy checking satisfiability of open transitions in the course of the residual algorithm, and show that  the reachable part of the open automaton can be generated and it is minimal with respect to the decidability of predicate satisfiability.
\item We show the usage of this approach on various use-cases from different settings, including an
    industry-inspired case-study, namely one architectural pattern
    extracted (and extended) from the BIP specification of a nanosatellite on-board
    software.
\end{itemize}

@@

%% There is not much research work towards trying 
Very few attempts were made to develop symbolic bisimulation approaches
for the value-passing process algebras and languages\mdash our long-term goals\mdash
especially, there is no algorithmic treatment of the symbolic systems developed by
interacting with automatic theorem provers. The closest work is the
one already mentioned by Rathke~\cite{HennessyRathke:TCS98},
who developed the symbolic bisimulation for a
calculus of broadcasting system (CBS). CBS is similar to classic
process calculi such as CCS and CSP, but communicating by broadcasting
values, %one-to-many communication instead of one-to-one communication and
transmitting values without blocking. That makes the definition of
the symbolic semantics and bisimulation equivalence different from the
classic works.

For other applications, such as the analysis of 
programming languages, there exist dedicated platforms using
external automatic theorem 
provers (ATP) or automatic tactics from interactive theorem
provers (ITP), to perform symbolic reasoning, and for example to
discharge some subgoals in the proofs.
Tools like Rodin~\cite{deharbe2013,deharbe2014} have
already integrated several provers, like Z3, as modules for proving
the proof obligations generated from a user model. 
The prover we use, which also happens to be Z3, is developed by Microsoft Research
based on the satisfiability modulo 
theories framework (SMT), and is mainly applied in extended static checking, test case
generation, and predicate abstraction.
In a similar way, there are several ATPs/ITPs we could consider to use for
result pruning and bisimulation checking in our algorithm, as an
alternative to Z3, such as CVC4~\cite{barrett:CAV2011},
Coq~\cite{armand:CPP2011} or others.

@@

\paragraph{Structure.}
In Section
\ref{section:pnets} we give a description and a formal definition of
the pNet model. In section
In Section \ref{section:op-semantics} we briefly recall  the operational semantics
of pNets.
In Section \ref{section:implementation} we present an algorithm to compute this semantics, including the interaction with Z3.
In Section~\ref{section:examples} we give a few use-cases, in particular, a BIP
architecture from the nano-satellite case-study. 
Finally we conclude and discuss perspectives in Section
\ref{section:conclusion}.

@@

We first introduce pNets and some notations. Then we give the formal definition of pNet structures, together with an operational semantics for open pNets.

pNets are tree-like structures, where the leaves are either
\emph{parameterized labelled transition systems (PLTSs)}, expressing the
behaviour of basic processes, or \emph{holes}, used as placeholders
for unknown processes, of which we only specify their set of possible
actions, named \emph{sort}.
Nodes of the tree (pNet nodes) are synchronizing artifacts, using a
set of \emph{synchronization vectors} that express the possible
synchronization between the parameterized actions of a subset of the
sub-trees.

%\smallskip\noindent
\paragraph*{Notations.} \
We extensively use indexed structures
over some countable indexed sets, which are equivalent to mappings over
the countable set. % . The indexes will usually be
% integers, bounded or not. Such an indexed family is
%denoted
%follows:
The symbole $a_i^{i\in I}$
denotes a family of elements $a_i$ indexed over the
set $I$.
When this is unambiguous, we shall use notations for sets, and typically write
``indexed set over I'' which formally means multisets, and
write $x\in a_i^{i\in I}$ to mean $\exists i\in I.\, x=a_i$.  An empty
family is denoted $\emptyset$. We
denote classically $\overline{a}$ a family when the indexing set is
irrelevant.  $\uplus$ is the disjoint union on indexed sets.

\paragraph*{Term algebra.} \
Our models rely on a notion of parameterized actions that are
symbolic expressions using data types and variables. As we want to encode
the low-level behaviour of possibly very different
programming languages, we do not want to impose one specific algebra
for denoting actions, nor any specific communication mechanism. So we
leave unspecified the constructors of the algebra that will allow building
expressions and actions. Moreover, we use a generic {\em action interaction}
mechanism, based on unification between two or more action
expressions. This will be used in the semantics of synchronization
vectors to express various kinds of communication or synchronization mechanisms.

@@

A PLTS is a labelled transition system with variables, which can be
manipulated, defined, or accessed inside states, actions, guards, and
assignments. 
%
%Without loss of generality and to simplify the formalization, we suppose 
%here that variables are local to 
%state: each state has its set of variables disjoint from the others. Transmitting 
%variable values from one state to the other can be done by explicit assignment. 
%
Each state has its set of variables called \emph{state variables}, 
which can only be modified by the assignment in transitions targeting this state. 
A global state variable of a PLTS is a state variable defined in all states.
% $x$ satisfying $\forall s\!\in\! S .\, x\!\in\! \vars(s)$. 
%
%Similarly, to simplify the management of variables and without loss of expressivity, we 
%suppose that transitions looping to the same state does not do assignments.
Note that we make no assumptions on finiteness of the set of states, nor
on finite branching of the transition relation.

We first define the set of actions a PLTS can use.  Let $a$
range over action labels, $\symb{op}$ are operators, and $x_i$ range over
variable names. Action terms are:
\[
\begin{array}[l]{rcl@{\quad}p{5.5cm}}
  \alpha\in\AlgA&::=&a(p_1,\ldots,p_n)&\text{action terms}\\
  p_i&::=& ~\symb{Expr}&\text{parameters}\\
  \symb{Expr}&::=& \symb{Value}~|~x~|~\symb{op}(\symb{Expr}_1,\dots,\symb{Expr}_n)&\text{expressions}
\end{array}
\]
%\[
%\begin{array}[l]{rcl@{\quad}p{5.5cm}}
%  \alpha\in\AlgA&::=&a(p_1,\ldots,p_n)&\text{action terms}\\
%  p_i&::=& ?x~|~\symb{Expr}&\text{parameters (input variable or expression)}\\
%  \symb{Expr}&::=& \symb{Value}~|~x~|~\symb{op}(\symb{Expr}_1,..,\symb{Expr}_n)&\text{Expressions}
%\end{array}
%\]

\begin{definition}[PLTS]
\label{PLTS}
%\noteSB{It's a bit strange that algebra presentations are not referred to here}
Given a term algebra $\Talg$, a PLTS is a tuple
$PLTS\triangleq\mylangle S,s_0, \to\myrangle$ where:
\begin{itemize}
\item[$\bullet$]
$S$ is a set of \emph{states}, with $s_0 \in S$ the \emph{initial state}.
\item[$\bullet$] $\to \subseteq S \times L \times S$ is the \emph{transition relation}, with
$L$ the set of labels of the form
$\langle \alpha,~e_b,~(x_j\!:= {e}_j)^{j\in J}\rangle$,
  where $\alpha \in\AlgA_{\P}$ is a parameterized action,
    $e_b \in \AlgB_{\P}$ is a guard, and
    expressions  $\AlgE_{\P}\cup\AlgA_{P}$ are assigned to $x_j$.
If 
$s \xrightarrow{\langle \alpha,~e_b,~(x_j\!:= {e}_j)^{j\in
		J}\rangle} s'$ then 
                $\vars(e_b)\!\subseteq\! \vars(s)\cup\vars(\alpha)$, and
		$\forall j\!\in\! J .\,\vars(e_j)\!\subseteq\! \vars(s)\land 
x_j\!\in\!\vars(s')$.
%\item[$\bullet$]
%The $x_j$ are \emph{state variables} of %% the state $s'$. 
\end{itemize}
\end{definition}

Now, we define
pNet nodes as constructors for hierarchical structures.
A pNet node has a set of sub-pNets that can be either pNets or PLTSs, and a
set of \emph{holes}, playing the role of process parameters
(i.e. unknown in the environment).

@@

%% \begin{definition}[pNets]\label{def-pnets}
%% A pNet is a hierarchical structure, whereof the leaves are PLTSs and holes:
%% $\pNet::= PLTS~|~\mylangle \pNet_i^{i\in I}, S_j^{j\in J}, \symb{SV}_k^{k\in K}\myrangle$,
%% where
%% \begin{itemize}
%% \item[$\bullet$] %\noteEM{Pire que ca, $\I$ n'est pas
%%                  %defini...}\noteSB{Why is one $\I$ and the other
%%   %$\I_\P$?}
%%   $I$ and $J$ are the sets over which are indexed sub-pNets and holes, respectively, such that: $I\cap J=\emptyset$ and  $I\cup J\neq\emptyset$.

@@

%% \item[$\bullet$] $S_j \subseteq \AlgA$ is a set of action terms,
%%   denoting the $\Sort$\footnote{The definition of $Sorts$ (the
%%     set of actions of a Hole  
%%   or pNet), and also of $Leaves$ and $Holes$ (respectively all PLTSs
%%   and all holes in a pNet), can be found in \cite{henrio:Forte2016}.}
%%   of hole $j$.

@@

  \begin{definition}[pNets]\label{def-pnets}
    A pNet is a hierarchical structure where leaves are PLTSs and holes:
   %%  \footnote{Typing can 
%% be added to pNets by defining and checking their interfaces, called 
%% sort~\cite{HMZ-FORTE2016}}
$\pNet\triangleq PLTS~|~\mylangle \pNet_i^{i\in I}, J, \symb{SV}_k^{k\in 
K}\myrangle$, with $I, J, K$ potentialy infinite,
where
\begin{itemize}
\item[$\bullet$] $\pNet_i^{i\in I}$ is the family of sub-pNets;
%  $\pNet_i^{i\in I}$ is a family of sub-pNets where $I\in\I_\P$ is the set over which 
%sub-pNets are indexed.

@@

In Figure~\ref{schema:EnableEncodings}, we show examples of these
constructs, with two PLTSs, one hole and one pNet node encoding our running example.

@@

The semantics of open pNets will be defined as an open automaton, that
is an automaton where each transition composes transitions 
of several
% \noteSB{These should all be PLTSs, right? Or did I miss
%  something?}\noteEM{Hum, nor really, something intermediate; the actions there can be
%  parameterised actions, but the label has no guard and no
%  assignments}
LTSs with the actions of some holes; the transition occurs if some predicates hold,
and can involve a set of state modifications. Each state of an open
automaton has a set of \emph{state variables} that can be assigned by
incoming transitions. Strictly speaking, the LTSs at the leaves of the
open automaton are a restricted form of PLTSs, where labels are
parametrised actions, but include no guard nor assignments.

@@

%
%Then the semantics of a pNet is characterized by a set of {\em open
%transitions}, where the hypotheses on process parameters are
%replaced by 1) transitions of the PLTSs at the leaves, and 2) formal
%hypotheses on the transitions of the holes. A {\em predicate} is used
%to relate the parameters and names appearing in the actions of the
%leaves and the holes involved in the rules, but also appearing in  the resulting action.

@@

\paragraph{States of open pNets.}\label{def-states}
  A state of an open pNet is a tuple of the
  states of its leaves (in which we denote tuples
  in structured states as $\triangleleft\ldots\triangleright$).
  For any pNet p, let $\overline{\mathit{Leaves}} = \mylangle S_i,{s_i}_0, \to_i\myrangle^{i \in L}$ be the set of PLTS at its leaves,
  then $\mathit{States}(p) = \{\ostate{s_i^{i\in L}}
  |\, \forall i\in L. s_i \in S_i\}$.
A PLTS being its own single leave:
  $\mathit{States}(\mylangle S,s_0, \to\myrangle) = \{\ostate{s}|\, s \in S\}$.
The initial state is defined as:
$\mathit{InitState}(p) = \ostate{{{s_i}_0}^{i\in L}}$.

@@

\paragraph{Structural Semantic Rules:}
The semantics of pNet in term of open automata has been defined in
\cite{henrio:Forte2016}, in the form of 2 structural rules, one for
PLTSs, one for pNet nodes. These rules are slightly improved, adding guards in the synchronisation vectors and syntax for universal quantifier in the guards (see~\cite{Avocs-RR}).

In \cite{henrio:Forte2016} we also proved:
\begin{theorem}[Finiteness]
  \label{Th:finiteness}
  Let $pnet=\mylangle \overline{\pNet},\overline{S}, \symb{SV}_k^{k\in K}\myrangle$ be an open pNet  with leaves $l_i^{i\in I}$ and holes $h_j^{j\in
  J}$, where the sets $I$ and $J$ are finite. Suppose the synchronisation vectors of all pNets 
  included in  $pnet$   are finite, and 
% $\forall i \in L.\, finite{(states(l_i))} \text{ and } l_i$ 
$l_i $ has a finite number of state variables for each $i\in I$, then the semantics of $pnet$ is an open automaton 
%$\mathcal{T}$ 
with finitely many states and transitions.
\end{theorem}
  Remark that all the elements of such pNets and open automata being
symbolic, they can represent many classes of unbounded systems.
